---
title: "Analyzing Keystrokes"
author: "Gary Feng <gary.feng@gmail.com>"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Keystroke Analysis in Educational Tests #

To be expanded at a later time.

# Capturing keystrokes # 

Keystrokes are a misnomer. For our purposes, we track the changes in the text content using a 
diff algorithm. The text diff may be triggered by keyboard input, but also other editing actions
such as copy-and-paste, spell-checker use, etc.

*To be expanded*

# Data Structure and Transformation #

The following is an example of a "text.change" entry, where the letter 's' at position 51 was
removed. 

```
<observableDatum>
  <sceneId>AnswerQuestions</sceneId>
  <controlId>AnswerQuestions.TextArea</controlId>
  <eventType>text.change</eventType>
  <timestamp>2015-07-27T17:57:10.349Z</timestamp>
  <content>
    <pair>
      <key>position</key>
      <value>51</value>
    </pair>
    <pair>
      <key>removedText</key>
      <value> <![CDATA[ s ]]></value>
    </pair>
    <pair>
      <key>addedText</key>
      <value> <![CDATA[ ]]> </value>
    </pair>
  </content>
</observableDatum>
```

The ```text.change``` events are sandwitched between the ```text.focus``` and the ```text.blur``` events, which mark whether the text input area is activated or deactivated (e.g., when 
the user clicked
in and out of the box). We also need to distinghish text input on different scenes, which is
why we keep the ```nav.scene``` event as an input to the ```events2df()``` function. It is in the 
```dropEvents``` list, meaning that ```events2df``` will not output this event.

For analysis, we extract all the ```text.\*``` nodes into a data frame. Note the column 
```content``` which contains lists of attributes such as ```addedText``` in the XML above. This gives us the flexibility to store arbitrary information for different events without polluting the data
frame structure. We can access these attributes using the ```@``` operator defined in the 
[dfat](https://github.com/garyfeng/dfat) library. This hopefully achieves a balance between 
representational flexibility and the convenience of the data frame structure required by most
R statistical functions.

```{r, eval=FALSE}
require(devtools)
devtools::install_github('garyfeng/dfat')   # overwrites the @ sign
devtools::install_github('garyfeng/pdata')  # utilities for process data 

# read the data from all XML files under "./data"; 
# the filename (without extention) will be in the variable "bookletId"
data  <- readXML(Sys.glob("data/*.XML"), subjIdVar="bookletId")

# in this case the data frame will have the following variables
# bookletId: subj ID
# blockId: the condition the subj is in.
# sceneId: the scence in the test
# eventType: the event id, for each row
# ts: the timestamp in POSIXct
# content: a list with named members representing extended attributes; see above XML for an example
#  The <key> becomes the attribute name, and <value> becomes the value.

# get only events starting with "text"
keystrokes<- subset(data, grepl("^text", data$eventType))

```

----

We then analyze the inter-key intervals (IKIs) for various kinds of keystroke events, namely

* all keystrokes
* spaces, which most typists type automatically after a word
* delets, which would require some meta-cognitive monitoring and cognitive inhibition.
* inter-word, which are the iki before a word initial, defined as after a space but not a delet event
* intra-word, which are all non-inter-word alphabetic keys

We will also calculate a bunch of other useful variables.

``` {r, eval=FALSE}
# calc the durSinceTextFocus
t0<- keystrokes$ts; t0[keystrokes$eventType!="text.focus"]<-NA; t0[1]<-keystrokes$ts[1]
keystrokes$durSinceTextFocus <- as.numeric(timeDiff(t0, keystrokes$ts))

# note that position was a char var by default. Turning to numeric.
# the +1 has to do with the semantics of the cursor position; 
# At least with the keystroke capturing system we used, the textPos is the position at
# which the action is ABOUT TO HAPPEN, whereas the textLen is the lengh of the text
# after the editing action. We add 1 here to make them more or less consistent, although
# it does not solve the issue completely.
# TO-DO: think through the relation between  textPos and textLen
keystrokes$textPos <- as.numeric(keystrokes$content@"position") +1

# Before calculating the textLen variable, we first get the length of the change
# + for addedText, - for removedText; the editLen is the net difference.
# this one may be better left without using @
keystrokes$editLen <-sapply(keystrokes$content, function(c) {
  ifelse( (is.null(c$position)), 0, nchar(c$addedText)-nchar(c$removedText))
})

# add editLen over time (but within each user and textbox)
keystrokes %>% group_by(bookletId, controlId) %>% 
  mutate(textLen = cumsum(editLen)) -> keystrokes
keystrokes$textLen[is.na(keystrokes$textPos)] <-NA

# Classify keystrokes 
# TO-DO: very simple algorithm; this needs to be more sophisticated.
addedText <-keystrokes$content@"addedText"
allKeys <- which(! is.na(addedText))
# words are alphabetic
alphabetKeys<- which(addedText %in% c(letters, LETTERS))
# no addition, it means deletion; @@ we may miss the type-over event here
del <- which(addedText=="")   
# positions of spaces, the next ones are the word initials, let's assume
sp <- which(addedText==" ")   
# word initial, after space but the next key is not deleting
inter <- intersect(sp+1, alphabetKeys)  
# all alphabetic keys that are not inter are "intra"
intra <- setdiff(alphabetKeys, inter)  
# classifiation
keystrokes$keyType <-NA
keystrokes$keyType[sp] <-"Spaces"
keystrokes$keyType[del] <-"Delecting"
keystrokes$keyType[intra] <-"Intra-word"
keystrokes$keyType[inter] <-"Inter-word"

# prevent reordering in plotting
# TO-DO: this gives warnings of duplicated levels. What we want is that the levels are ordered
#   in the order of how they occurred in the data. There's got to be a better way to do this.
#   Elsewhere I manually set the order with levels=c("first", "second", ...)
keystrokes$sceneId <- factor(keystrokes$sceneId, levels = keystrokes$sceneId)
keystrokes$eventType <- factor(keystrokes$eventType, levels = keystrokes$eventType)

```

----

## Text Length and Cursor Position by Scenes and by Condition ##

The "ribbon" plot shows the progression of 2 variables over time

* the text length
* the cursor position

The difference between the two lines -- whenever the test-taker edit at positions other than the end of the text -- is shown in the shaded area. In other words, the "flags" signal non-local editing actions. 

In the following example, we plot the ribbon plot for each ```sceneId```, with time zero starting from
the onset of each scene (one textbox per scene at most). We also want to compare across a categorical 
variable ```blockId```, and identify each student (```bookletId```) with color.

```{r, eval=FALSE}
# stats of keystrokes
ggplot()+
  geom_ribbon(data=keystrokes, alpha=0.5, 
      aes(x=as.numeric(durSinceSceneStart)/60, ymin=as.numeric(textPos), ymax=as.numeric(textLen), 
          color=bookletId, fill=bookletId))+
  # titles and options
  xlab("Task Time (in min.)") +ylab("Text Length - Cursor Position")+ 
  theme(legend.position="none")+
  facet_grid(blockId ~ sceneId , scales="free_x")

```

----

Let's say we want to show the distribution of the inter-key-intervals (IKIs) for
different types of keystrokes, we can use the getDistr() function (part of the pdata library)
to compute distribution functions, get several distributions under a single data frame,
and plot them using ggplot.

``` {r, eval=FALSE}
# showing keystroke iki hazard and pdf
# we need "getDistr()" which in turn requires the muhaz library

binwidth = 25
# We want to get distribution functions for different conditions in a single data frame
# to make it easier to plot in ggplot2.
# getDistr() doesn't work well with dplyr's group_by() syntax for whatever reason.
# using the good-o for loop.
tmp <- getDistr(keystrokes$dur*1000, binwidth = binwidth) %>% mutate(keyType="All")
for (i in unique(keystrokes$keyType)) {
  if (is.na(i)) next 
  getDistr(keystrokes$dur[keystrokes$keyType==i] *1000, binwidth = binwidth) %>% 
    mutate(keyType=i) %>% 
    rbind(tmp) -> tmp
}


###############
# 4 panel graph

for (i in c("All", "Intra-word", "Inter-word", "Spaces", "Delecting")) {
  if (is.na(i)) next 
  pdfLog<- ggplot(data=tmp[tmp$keyType==i,], aes(x=time, y=pdf, color=sceneId))+
    geom_line() +
    labs(list(x = "InterKey Interval (msec)", y = "Prob. Density"))+
    xlim(0.001, 1000)
  
  pdfLogLog<- ggplot(data=tmp[tmp$keyType==i,], aes(x=time, y=pdf, color=sceneId))+
    geom_line() +
    labs(list(x = "InterKey Interval (msec)", y = "Prob. Density"))+
    #stat_smooth(se=F) +
    scale_x_log10() +scale_y_log10()+
    theme(legend.position="none")
  
  hazLog<- ggplot(data=tmp[tmp$keyType==i,], aes(x=time, y=haz, color=sceneId))+
    geom_line() +
    labs(list(x = "InterKey Interval (msec)", y = "Hazard Rate"))+
    xlim(0.001, 1500)+
    theme(legend.position="none")
  
  hazLogLog<- ggplot(data=tmp[tmp$keyType==i,], aes(x=time, y=haz, color=sceneId))+
    geom_line() +
    labs(list(x = "InterKey Interval (msec)", y = "Hazard Rate"))+
    #stat_smooth(se=F) +
    scale_x_log10() +scale_y_log10()+ theme(legend.position="none")
  
  grid.arrange(pdfLog, hazLog, pdfLogLog, hazLogLog, ncol=2, 
               top=textGrob(paste("KeyType = ", i), gp=gpar(cex=2)))
}
```

